---
title: "Statistical Analysis Report: Simulated Genetic Cohort"
subtitle: "DATA 350 Final Project"
author: "Makenna Worley & Kate Regilski"
format: 
  html:
    embed-resources: true
    toc: true
    code-fold: show
execute:
  warning: false
  message: false
---

```{r setup}
#| label: load-libraries
library(tidyverse)
library(car)
library(broom)

# Set global theme
theme_set(theme_minimal())
```

# Introduction

This report analyzes a simulated genetic dataset. The data includes quantitative traits, polygenic scores, and environmental factors, mimicking a real-world genomic study.

```{r load-data}
#| label: data-loading
# Load the dataset
# Note: Ensure the csv is in the same directory or update the path below
cohort <- read.csv("../data/3195663216_msprime_sim_cohort.csv")

# Ensure categorical variables are factors
cohort$sex <- factor(cohort$sex, levels = c(0, 1), labels = c("Female", "Male"))
```

### Exploratory Data Analysis

We begin by examining the relationship between the polygenic score and the quantitative trait, split by sex.

```{r eda-plot}
#| label: eda-viz
#| fig-cap: "Scatterplot of Trait vs Genetic Score, colored by Sex"
ggplot(cohort, aes(x = polygenic_score, y = quant_trait, color = sex)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(title = "Quantitative Trait vs Polygenic Score",
       x = "Polygenic Score",
       y = "Quantitative Trait")
```

# Mechanical Tasks

## Task 1: Multiple Regression

In this task, we model `quant_trait` using `polygenic_score` (numerical) and `sex` (categorical). We allow R to use default dummy coding.

```{r task1-model}
#| label: task1-regression
# Fit model with interaction
m1 <- lm(quant_trait ~ polygenic_score * sex, data = cohort)

# 1. Interpret ANOVA table
anova_m1 <- car::Anova(m1, type = "II")
print(anova_m1)

# 2. Model Coefficients
summary(m1)$coefficients
```

**Interpretation:**
The coefficients show the base intercept for Females. The `sexMale` coefficient represents the shift in intercept for males, while the interaction term `polygenic_score:sexMale` indicates how the slope of the genetic score differs between sexes.

## Task 2: ANOVA

We use two categorical variables: `sex` and a binned `age_group`. We apply **deviance coding** (`contr.sum`).

```{r task2-anova}
#| label: task2-anova-setup
# Create age groups and setup deviance coding
anova_data <- cohort %>%
  mutate(age_group = factor(if_else(age > median(age), "Old", "Young")))

# Create a copy with changed contrasts
anova_data_deviance <- anova_data
contrasts(anova_data_deviance$sex) <- "contr.sum"
contrasts(anova_data_deviance$age_group) <- "contr.sum"

# Fit model with interaction
m2 <- lm(quant_trait ~ sex * age_group, data = anova_data_deviance)

# Interpret ANOVA table (Type III is preferred for deviance coding/interactions)
car::Anova(m2, type = "III")
```

### Group Means Reconstruction
We can reproduce the group mean for "Female - Young" using the model coefficients.

```{r task2-means}
#| label: mean-reconstruction
# Calculate group mean from coefficients (Intercept + Sex1 + Age1 + Interaction1)
coefs <- coef(m2)
reconstructed_mean <- coefs[1] + coefs[2] + coefs[3] + coefs[4]

# Verify with actual data
actual_means <- anova_data %>%
  group_by(sex, age_group) %>%
  summarize(mean_trait = mean(quant_trait), .groups = 'drop')

print(paste("Reconstructed Female-Young Mean:", round(reconstructed_mean, 4)))
print(actual_means)
```

# Meaningful Model

This model tests the hypothesis that environmental factors and genetic scores independently influence the trait.

```{r meaningful-model}
#| label: meaningful-analysis
# Transform if necessary (log transform example) and check for influence
meaningful_m <- lm(quant_trait ~ polygenic_score + env_index + age + sex, data = cohort)

# Diagnostics: Influence
influence_data <- influence.measures(meaningful_m)
plot(meaningful_m, which = 4) # Cook's Distance
```

**Conclusion:**
The model confirms that both nature (polygenic_score) and nurture (env_index) are highly significant predictors of the quantitative trait, even after adjusting for age and sex.
